Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-30 curvature.tex
more graph algorithm explanation and citations

Thus although the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
\cite{Ding2011-bj} showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
$$ n - 2\ceil{\sqrt{n}} + 1
\le \Delta_{\text{uSPR}}(n)
\le n - 3 - \floorV{\frac{\sqrt{n - 2} - 1}{2}}.
$$
We are not aware of any other work investigating large-scale properties of the SPR graph.

@@

To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by simply successively adding trees to the graph:
\begin{enumerate}[label={\arabic*}.]
	\item Let $G$ be an empty graph
	\item For each of the $m$ trees in a specified but arbitrary order,
		\begin{enumerate}
			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index
			\item For each of the $\OhOf{n^2}$ neighbors of $T$
				\begin{enumerate}
					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index
				\end{enumerate}
		\end{enumerate}
\end{enumerate}

\begin{theorem}
	\label{thm:construct_graph}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	We implement the graph as an adjacency list with integer-labelled vertices.
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.\todo{"in ascending order to the end of the vertex list": confused!}
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$)\todo{sorry, but I don't understand this $\log n$ term from the explanation}.
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}
This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.

@@

\begin{lemma}
	\label{lem:select_random_neighbor}
	Given a tree $T$, an rSPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
\end{lemma}
\begin{proof}
	We apply the above procedure.
	We use a standard nodes and pointer representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
	Both of these traversals require $\OhOf{n}$-time.
	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
	Thus, the for loop requires linear time.
	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
\end{proof}

>>>>

Although the SPR graph is thus very important in determining the degree of success of phylogenetic inference procedures, still little is known about the rooted or unrooted versions of the SPR graph itself, or random walks thereupon.
\cite{Song2003-gf} developed a recursive procedure on a tree to find the degree of the corresponding vertex in the rooted SPR (rSPR) graph, and corresponding bounds on degree.
\cite{Ding2011-bj} showed that the diameter $\Delta_{\text{rSPR}}$ of the rSPR graph is $n - \Theta(\sqrt n)$, and for the unrooted case they show
$$ n - 2\ceil{\sqrt{n}} + 1
\le \Delta_{\text{uSPR}}(n)
\le n - 3 - \floorV{\frac{\sqrt{n - 2} - 1}{2}}.
$$
We are not aware of any other work investigating large-scale properties of the SPR graph.

@@

To quickly compute dense rSPR graphs (those containing a significant portion of the full rSPR graph) we can avoid pairwise computations by successively adding trees to the graph and efficiently identifying their neighbors:
\begin{enumerate}[label={\arabic*}.]
	\item Let $G$ be an empty graph
	\item For each of the $m$ trees in a specified but arbitrary order,
		\begin{enumerate}
			\item Add a vertex $i$ to the graph representing the current tree $T$ where $i$ is $T$'s order index
			\item For each of the $\OhOf{n^2}$ neighbors of $T$
				\begin{enumerate}
					\item If the current neighbor $S$'s index is in $G$ then add an edge $(j,i)$ to the graph, where $j$ is $S$'s order index
				\end{enumerate}
		\end{enumerate}
\end{enumerate}

\begin{theorem}
	\label{thm:construct_graph}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	We implement the graph with an adjacency list representation with integer-labelled vertices that supports $\OhOf{\log n}$ edge insertions and lookups (with e.g. red-black trees~\cite{guibas1978dichromatic}, as the vertex degrees are $\OhOf{n^2}$).
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list, which can be stored as a fixed-size array.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time.
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie~\cite{fredkin1960trie} using the $\OhOf{n}$-length Newick \cite{wiki:newick} representation of the tree as a string.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes-and-pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to phylogenetic trees with string labels).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}
This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}~\cite{spr_neighbors}, which outputs an edge list format graph suitable for input to other software.
The construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR graph from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon rSPR graph in this paper, our algorithm required only 303.45 seconds to construct it on the same hardware.
Constructing the 8-taxon rSPR graph using the previous method required 377,395 seconds (more than 4 days), and thus is infeasible for constructing larger rSPR tree graphs.
Thus, we believe this fast graph construction procedure will itself be useful for further studies of rSPR graph subsets similar to~\cite{Whidden2015-yi}.

@@

\begin{lemma}
	\label{lem:select_random_neighbor}
	Given a tree $T$, an rSPR neighbor of $T$ can be chosen uniformly at random in $\OhOf{n}$-time using $\OhOf{n}$ space.
\end{lemma}
\begin{proof}
	We apply the above procedure.
	We use a standard nodes-and-pointers representation of the trees, which can be constructed in $\OhOf{n}$-time from a newick string representation and uses linear space in $n$.
	We can compute the degree of $T$ in linear time and space using Lemma~\ref{lem:compute_degree}.
	To efficiently compute $\size{N(T,u)}$ for each node $u$ of $T$, we require the number of nodes $x$ in the subtree rooted at $u$.
	We precompute these by (1) labeling each node with its preorder number in a preorder traversal and (2) summing the number of descendant nodes in a postorder traversal and storing the results in an array indexed by preorder number.
	Both of these traversals require $\OhOf{n}$-time.
	There are $2n-1$ = $\OhOf{n}$ nodes of $T$, and $\size{N(T,u)}$ can be computed in constant time using the subtree sizes.
	Moreover, the tree $S$ can be found in $\OhOf{n}$-time by iterating over the edges of $T$ that are not contained within $u$'s subtree to select the corresponding rSPR destination.
	Finally, we require linear time to apply the chosen rSPR operation which entails removing a node, adding a node, and updating a constant number of pointers.
	Thus, the for loop requires linear time.
	By Lemma~\ref{lem:compute_degree} the chosen tree is an rSPR neighbor of $T$ and is chosen uniformly at random.
	Therefore, the procedure uses linear time and space and selects an rSPR neighbor of $T$ uniformly at random.
\end{proof}
